---
title: "Assignment 1 - Setting Up R"
author: "Elizabeth Crawley"
date: "2025-05-31"
output: word_document
---

## 1. Install the ISLR library using the install.packages() command.
## Call the library using the library(ISLR) command to ensure that the library is correctly installed 

```{r}
library(ISLR)
```

## 2. Create a new R-Notebook (.Rmd) file. 
## In the first code chunk, call the ISLR library and then print the summary of the Carseats dataset. 
## How many observations (rows) this dataset contains?
## I used the summary() function to view the carseats dataset.
```{r}
summary(Carseats)
```

## I used rnow() function to view the number of rows.

```{r}
nrow(Carseats)
```

## 3. Using the summary statistics shown above, what is maximum value of the advertising attribute?
## I used the summary() function to view the Advertising attribute of the CarSeats dataset.

```{r}
summary(Carseats$Advertising)
```

## 4. Calculate the IQR of the Price attribute
## I did not know what IQR stood for, so I had to search online to learn I could use the IQR() function to view the "Interquartile Range" which is the difference between the 75th percentile and the 25th percentile, IQR = Q3-Q1. 

```{r}
IQR(Carseats$Price)
```

##5. Plot the Sales against Price. 
## What do you see in there? Calculate the correlation of the two attributes. 
## What does the sign of the correlation coefficient suggest?

```{r}
plot(Carseats$Price, Carseats$Sales)
```

## The plot could be cleaned up by adding a descriptive title to the x and y axis, choosing a different style of plot such as a box plot, or making the plot dots solid or blue, for example. 
## Without adding these, it is difficult to determine the correlation coefficient. 
## However, I found online that I could use the cor() function to find the correlation coeffient. 
```{r}
cor(Carseats$Price, Carseats$Sales)
```

## The correlation coefficient can be calculated using the cor() function in R. ## The correlation coefficient is approximately -0.44
## The sign (-) of the correlation coefficient means that as the price increases, sales tend to decrease. 
## Being unfamiliar with what a negative sign would indicate, I searched online and found the answers I provided above.

