---
title: "Churn Interpretative Analysis"
author: "Elizabeth Crawley"
date: "`r Sys.Date()`"
output: pdf_document
---
## Dataset Overview

The dataset "Churn", renamed as "my_data" consists of 1000 rows and 9 columns.
The columns are described as follows:

- CallFailures: Number of call failures.
- SubscriptionLength: Length of subscription in months.
- DataUsage: Amount of data usage.
- VoiceMinutes: Number of voice minutes used.
- CustomerSupportCalls: Number of calls to customer support.
- ContractType: Type of contract (Monthly or Annual).
- MonthlyCharges: Monthly charges.
- RoamingUsage: Amount of roaming usage.
- Churn: Binary indicator of whether the customer churned (1) or not (0).
```{r}
my_data <- read.csv("C:\\Users\\ecrawle1\\OneDrive - Kent State University\\Desktop\\Machine Learning\\Churn.csv")
head(my_data)
```
```{r}
Churn <- read.csv("C:/Users/ecrawle1/OneDrive - Kent State University/Desktop/Machine Learning/Churn.csv")
 View(Churn)
 tail(my_data)

```
```{r}
structure(my_data)
```
# Statistical summary for my_data
```{r}
summary(my_data)
```
## Statistical summary for CallFailures
```{r}
mean(my_data$CallFailures)
median(my_data$CallFailures)
table_my_data <- table(my_data$CallFailures)
mode_value <- as.numeric(names(table_my_data)[table_my_data == max(table_my_data)])
print(mode_value)
```
## Statistical summary for SubscriptionLength
```{r}
mean(my_data$SubscriptionLength)
median(my_data$SubscriptionLength)
table_sublength <- table(my_data$SubscriptionLength)
mode_sublength <- as.numeric(names(table_sublength)[table_sublength == max(table_sublength)])

```

## Statistical summary for DataUsage
```{r}
mean(my_data$DataUsage)
median(my_data$DataUsage)
mode_datausage <- max(my_data$DataUsage)
print(mode_datausage)
```

## Statistical summary for VoiceMinutes
```{r}
mean(my_data$VoiceMinutes)
median(my_data$VoiceMinutes)
mode_VoiceMinutes <- max(my_data$VoiceMinutes)
print(mode_VoiceMinutes)
```

## Statistical summary for CustomerSupportCalls
```{r}
mean(my_data$CustomerSupportCalls)
median(my_data$CustomerSupportCalls)
mode_CustomerSupportCalls <- max(my_data$CustomerSupportCalls)
print(mode_CustomerSupportCalls)
```
## Statistical summary for ContractType
## This is a categorical variable with values Monthly and Annual
 
## Statistical summary for MonthlyCharges
```{r}
mean(my_data$MonthlyCharges)
median(my_data$MonthlyCharges)
mode_MonthlyCharges <- max(my_data$MonthlyCharges)
print(mode_MonthlyCharges)
```
## Statistical summary for RoamingUsage
```{r}
mean(my_data$RoamingUsage)
median(my_data$RoamingUsage)
mode_RoamingUsage <- max(my_data$RoamingUsage)
print(mode_RoamingUsage)
```
## Statistical summary for Churn
```{r}
mean(my_data$Churn)
median(my_data$Churn)
mode_Churn <- max(my_data$Churn)
print(mode_Churn)
```
## There is no missing data
```{r}
total_missing <- sum(is.na(my_data))
total_missing
```
## The maxium SubscriptionLength across all customers is 24
```{r}
max(my_data$SubscriptionLength)
```
## Select and compare two or more variables from the dataset, such as Data Usage and CHURN. Create a table for the selected variables using the df(dataframe) and select commands.

```{r}

```
##Transform at least one variable (any transformation such as log(x), sqrt(x), scale(x), is acceptible) and generate a plot using ggplot for the fransformed variable.
```{r}
my_data <- read.csv("C:/Users/ecrawle1/OneDrive - Kent State University/Desktop/Machine Learning/Churn.csv")
selected_data <- my_data[, c("DataUsage", "SubscriptionLength")]
summary(selected_data)
library(dplyr)
data_summary <- selected_data %>%
  group_by(SubscriptionLength) %>%
  summarise(
    MeanDataUsage = mean(DataUsage, na.rm = TRUE),
    MedianDataUsage = median(DataUsage, na.rm = TRUE),
    SDDataUsage = sd(DataUsage, na.rm = TRUE)
  )
print(data_summary)
data_usage_by_sublength <- table(selected_data$SubscriptionLength, selected_data$DataUsage)
data_usage_by_sublength_df <- as.data.frame(data_usage_by_sublength)
print(data_usage_by_sublength_df)
library(ggplot2)
ggplot(selected_data, aes(x = factor(SubscriptionLength), y = DataUsage)) +
  geom_boxplot() +
  labs(x = "Churn Status", y = "Data Usage", title = "Data Usage by Subscription Length") +
  theme_minimal()

```

